---
format:
  html: 
    title: "Migration as System Stabilzer"
    author: "Sechang Kim"
    date: "05/27/2026"
    date-modified: last-modified
    toc: true
    execute:
      echo: false
categories:
  - Migration
  - Climate Migration
  - Spatial Analysis
  - GEOG 544
---

**Research question.** How are cumulative and recurrent flood-inundation damages spatially distributed across Seoul's administrative *dongs*, and how might these patterns relate to housing-market conditions and residential mobility?

This research notebook begins by preparing the flood-damage point records and visualizing them against Seoul's administrative boundary layers. The purpose of this first section is exploratory: to inspect the annual spatial distribution of reported flood-inundation damage before constructing dong-year summaries, recurrence measures, or cluster typologies.

## 1. Basic visualization and data preparation

The raw flood-damage data are stored as annual point shapefiles. The `seoul_EMD` and `seoul_SGG` shapefiles are polygon boundary layers used as map context: EMD boundaries provide the finer administrative-dong geography, while SGG boundaries provide the higher-level district geography. The flood-damage attributes are read from the annual point files only.

```{r}
#| label: load-packages
#| message: false
#| warning: false

# Core spatial and data-wrangling packages
library(sf)
library(dplyr)
library(purrr)

# Interactive mapping packages
library(leaflet)
library(leaflet.extras2)
library(htmltools)
library(patchwork)
library(ggplot2)
library(ggdendro)
library(tidyr)
library(tmap)
library(broom)
library(car)
library(kableExtra)
library(nnet)
```

```{r}
#| label: read-flood-and-boundary-data
#| message: false
#| warning: false

# Data are stored in the project-level data directory. The first path works
# when running interactively from the project root; the second works when
# Quarto executes this document from the posts directory.
data_dir <- if (dir.exists("./data/GEOG544")) {
  "./data/GEOG544"
} else {
  "../data/GEOG544"
}

# These are the raw flood-damage fields retained from each annual point file.
raw_flood_fields <- c(
  "ADM_CD",
  "F_ZONE_NM",
  "F_YR",
  "F_AREA",
  "F_SHIM",
  "F_AVR_HGT"
)

# Read administrative boundary layers.
# EMD is the lower-level administrative-dong boundary; SGG is the higher-level
# district boundary. These polygon layers are used here as map context.
seoul_emd <- st_read(
  file.path(data_dir, "seoul_EMD.shp"),
  quiet = TRUE
) |>
  st_make_valid()

seoul_sgg <- st_read(
  file.path(data_dir, "seoul_SGG.shp"),
  options = "ENCODING=CP949",
  quiet = TRUE
) |>
  st_make_valid() |>
  st_transform(st_crs(seoul_emd))

# Find all annual flood-damage point shapefiles that follow the seoul_YYYY.shp
# naming pattern. Years with no reported records, such as 2015 and 2021, are
# simply absent from this file list.
annual_flood_files <- list.files(
  data_dir,
  pattern = "^seoul_[0-9]{4}\\.shp$",
  full.names = TRUE
) |>
  sort()

# Read each annual point file, keep only the key raw fields, transform it to
# the EMD CRS, and combine the results into one long-form sf object.
flood_points <- annual_flood_files |>
  map(
    \(path) {
      st_read(path, options = "ENCODING=CP949", quiet = TRUE) |>
        select(all_of(raw_flood_fields)) |>
        mutate(ADM_CD = as.character(ADM_CD),
               F_AVR_HGT = as.double(F_AVR_HGT),
               F_YR = as.integer(F_YR)) |>
        st_transform(st_crs(seoul_emd))
    }
  ) |>
  bind_rows()
```

The resulting object, `flood_points`, is a long-form `sf` point layer that combines all available annual flood-damage records. It retains the original raw flood fields, including `ADM_CD`, `F_ZONE_NM`, `F_YR`, `F_AREA`, `F_SHIM`, and `F_AVG_HGT`. The EMD and SGG polygon layers are retained separately as contextual boundary layers for visualization.

For the interactive map, the substantive time unit is the reported flood year, `F_YR`. Rather than using `leaflet.extras2::addTimeslider()`, which can behave poorly when many points share the same timestamp, this map uses a small custom JavaScript slider with one step per reported year. Moving the slider filters the displayed points directly by `F_YR`.

```{r}
#| label: annual-flood-time-slider-map
#| message: false
#| warning: false

# Leaflet expects longitude-latitude coordinates, so all map layers are
# transformed to WGS 84 for web mapping.
seoul_emd_map <- seoul_emd |>
  st_transform(4326)

seoul_sgg_map <- seoul_sgg |>
  st_transform(4326)

flood_points_map <- flood_points |>
  st_transform(4326) |>
  mutate(
    F_YR = as.integer(F_YR),
    popup = paste0(
      "<strong>Year:</strong> ", htmlEscape(F_YR), "<br/>",
      "<strong>Location:</strong> ", htmlEscape(F_ZONE_NM), "<br/>",
      "<strong>Raw ADM_CD:</strong> ", htmlEscape(ADM_CD), "<br/>",
      "<strong>Inundated area:</strong> ", htmlEscape(F_AREA), " m<sup>2</sup><br/>",
      "<strong>Flood depth:</strong> ", htmlEscape(F_SHIM)
    )
  )

# Extract point coordinates and attributes for the custom browser-side slider.
# The slider has one step per unique F_YR and redraws only that year's points.
flood_coords <- st_coordinates(flood_points_map)

slider_points <- flood_points_map |>
  st_drop_geometry() |>
  mutate(
    lng = flood_coords[, "X"],
    lat = flood_coords[, "Y"]
  ) |>
  select(lng, lat, F_YR, popup)

slider_years <- sort(unique(slider_points$F_YR))

# Interactive exploratory map of reported flood-damage locations by year.
leaflet() |>
  addProviderTiles(providers$CartoDB.Positron) |>
  addPolygons(
    data = seoul_emd_map,
    color = "#333333",
    weight = 0.4,
    opacity = 0.6,
    fillOpacity = 0,
    group = "EMD boundaries"
  ) |>
  addPolygons(
    data = seoul_sgg_map,
    color = "#333333",
    weight = 1.2,
    opacity = 1,
    fillOpacity = 0,
    group = "SGG boundaries"
  ) |>
  addLayersControl(
    overlayGroups = c("EMD boundaries", "SGG boundaries"),
    options = layersControlOptions(collapsed = FALSE)
  ) |>
  htmlwidgets::onRender(
    "
    function(el, x, data) {
      var map = this;
      var years = data.years;
      var points = data.points;
      var markerLayer = L.layerGroup().addTo(map);
      var playTimer = null;

      var control = L.control({ position: 'bottomleft' });

      control.onAdd = function() {
        var div = L.DomUtil.create('div', 'year-slider-control');
        div.style.background = 'white';
        div.style.padding = '8px 10px';
        div.style.border = '1px solid #999';
        div.style.borderRadius = '4px';
        div.style.boxShadow = '0 1px 4px rgba(0, 0, 0, 0.25)';
        div.style.minWidth = '220px';
        div.innerHTML =
          '<div style=\"font-weight: 600; margin-bottom: 4px;\">Year: ' +
          '<span id=\"flood-year-label\">' + years[0] + '</span></div>' +
          '<input id=\"flood-year-slider\" type=\"range\" min=\"0\" max=\"' +
          (years.length - 1) + '\" step=\"1\" value=\"0\" style=\"width: 200px;\">' +
          '<button id=\"flood-year-play\" type=\"button\" style=\"display: block; margin-top: 6px; padding: 3px 8px;\">Play</button>';

        L.DomEvent.disableClickPropagation(div);
        L.DomEvent.disableScrollPropagation(div);
        return div;
      };

      control.addTo(map);

      var slider = document.getElementById('flood-year-slider');
      var label = document.getElementById('flood-year-label');
      var playButton = document.getElementById('flood-year-play');

      function drawYear(index) {
        var year = Number(years[index]);
        label.textContent = year;
        markerLayer.clearLayers();

        for (var i = 0; i < points.F_YR.length; i++) {
          if (Number(points.F_YR[i]) === year) {
            L.circleMarker([points.lat[i], points.lng[i]], {
              radius: 1,
              stroke: true,
              color: '#8b1e3f',
              weight: 1,
              fillColor: '#d95f02',
              fillOpacity: 0.75
            })
              .bindPopup(points.popup[i])
              .addTo(markerLayer);
          }
        }
      }

      function stopPlayback() {
        if (playTimer !== null) {
          clearInterval(playTimer);
          playTimer = null;
          playButton.textContent = 'Play';
        }
      }

      function startPlayback() {
        playButton.textContent = 'Stop';
        playTimer = setInterval(function() {
          var nextIndex = Number(slider.value) + 1;

          if (nextIndex >= years.length) {
            stopPlayback();
            return;
          }

          slider.value = nextIndex;
          drawYear(nextIndex);
        }, 1000);
      }

      slider.addEventListener('input', function(event) {
        stopPlayback();
        drawYear(Number(event.target.value));
      });

      playButton.addEventListener('click', function() {
        if (playTimer === null) {
          if (Number(slider.value) >= years.length - 1) {
            slider.value = 0;
            drawYear(0);
          }
          startPlayback();
        } else {
          stopPlayback();
        }
      });

      drawYear(0);
    }
    ",
    data = list(
      years = slider_years,
      points = slider_points
    )
  )
```

This map is intended as an initial diagnostic visualization. The lighter EMD boundaries show the fine-grained neighborhood geography, while the thicker SGG boundaries make the broader district structure easier to read. Later sections will use these point records to construct administrative-dong summaries and develop measures of cumulative damage and recurrence.

## 2. Administrative-dong flood indicators for clustering

The following indicators are dong-specific cumulative measures for later cluster analysis. The unit of analysis is the 2025 administrative dong, so flood records are first assigned to the fixed 2025 boundary framework with a spatial point-in-polygon join rather than an attribute join on historical `ADM_CD`. The final object has one row per dong and retains the 2025 dong geometry for mapping. These are not dong-year indicators: annual records are used only as inputs for cumulative, recurrence, and concentration summaries.

```{r}
#| label: create-dong-flood-indicators
#| message: false
#| warning: false

# Use the 2025 administrative-dong boundary as the fixed target geography.
# Flood points are assigned to this framework spatially, not by joining on the
# historical raw ADM_CD values stored in the flood files.
dong_boundary_2025 <- seoul_emd |>
  mutate(ADM_CD_2025 = as.character(ADM_CD))

# Ensure both spatial layers use the same CRS before point-in-polygon joining.
if (st_crs(flood_points) != st_crs(dong_boundary_2025)) {
  flood_points <- flood_points |>
    st_transform(st_crs(dong_boundary_2025))
}

# Preserve the original point-level administrative code and spatially attach
# the 2025 administrative-dong identifier from the polygon layer.
flood_points <- flood_points |>
  mutate(ADM_CD_raw = as.character(ADM_CD))

flood_points_2025 <- flood_points |>
  st_join(
    dong_boundary_2025 |>
      select(ADM_CD_2025),
    join = st_within,
    left = FALSE
  )

# Prepare the spatially joined flood records used to construct dong-level
# indicators. Years with no reported records, 2015 and 2021, are excluded.
flood_records_for_indicators <- flood_points_2025 |>
  st_drop_geometry() |>
  mutate(
    F_YR = as.integer(F_YR),
    F_AREA = coalesce(as.double(F_AREA), 0),
    F_SHIM = coalesce(as.double(F_SHIM), 0),
    area_depth = F_AREA * F_SHIM
  ) |>
  filter(
    F_YR >= 2010,
    F_YR <= 2025,
    !F_YR %in% c(2015, 2021)
  )

# Cumulative damage indicators: counts, total inundated area, and total
# area-depth are each summarized once per administrative dong.
dong_cumulative_damage <- flood_records_for_indicators |>
  group_by(ADM_CD_2025) |>
  summarise(
    total_damage_count = n(),
    total_F_AREA = sum(F_AREA, na.rm = TRUE),
    total_area_depth = sum(area_depth, na.rm = TRUE),
    n_damage_years = n_distinct(F_YR),
    .groups = "drop"
  )

# Annual area-depth is used only to measure temporal recurrence and
# concentration within each dong.
dong_annual_area_depth <- flood_records_for_indicators |>
  group_by(ADM_CD_2025, F_YR) |>
  summarise(
    annual_area_depth = sum(area_depth, na.rm = TRUE),
    .groups = "drop"
  )

dong_area_depth_concentration <- dong_annual_area_depth |>
  group_by(ADM_CD_2025) |>
  mutate(
    total_area_depth = sum(annual_area_depth, na.rm = TRUE),
    p_t = if_else(total_area_depth > 0, annual_area_depth / total_area_depth, 0)
  ) |>
  summarise(
    n_damage_years_for_evenness = n_distinct(F_YR),
    area_depth_entropy = -sum(if_else(p_t > 0, p_t * log(p_t), 0)),
    max_year_share_area_depth = if_else(
      first(total_area_depth) > 0,
      max(annual_area_depth, na.rm = TRUE) / first(total_area_depth),
      0
    ),
    .groups = "drop"
  ) |>
  mutate(
    area_depth_evenness = if_else(
      n_damage_years_for_evenness > 1,
      area_depth_entropy / log(n_damage_years_for_evenness),
      0
    )
  ) |>
  select(ADM_CD_2025, area_depth_evenness, max_year_share_area_depth)

# Join the summaries back to all 2025 administrative dongs. Dongs with no
# reported flood damage are retained and assigned zero values.
dong_flood_indicators <- dong_boundary_2025 |>
  left_join(dong_cumulative_damage, by = "ADM_CD_2025") |>
  left_join(dong_area_depth_concentration, by = "ADM_CD_2025") |>
  mutate(
    across(
      c(
        total_damage_count,
        total_F_AREA,
        total_area_depth,
        n_damage_years,
        area_depth_evenness,
        max_year_share_area_depth
      ),
      \(x) coalesce(x, 0)
    ),
    log_total_damage_count = log1p(total_damage_count),
    log_total_F_AREA = log1p(total_F_AREA),
    log_total_area_depth = log1p(total_area_depth)
  ) |>
  select(
    ADM_CD_2025,
    ADM_NM,
    SGG_CD,
    total_damage_count,
    log_total_damage_count,
    total_F_AREA,
    log_total_F_AREA,
    total_area_depth,
    log_total_area_depth,
    n_damage_years,
    area_depth_evenness,
    max_year_share_area_depth
  )

# Inspect the completed sf object before using it in any cluster analysis.
dong_flood_indicators |>
  st_drop_geometry() |>
  summarise(
    n_dongs = n(),
    dongs_with_damage = sum(total_damage_count > 0),
    max_damage_count = max(total_damage_count),
    max_total_F_AREA = max(total_F_AREA),
    max_total_area_depth = max(total_area_depth)
  )

dong_flood_indicators |>
  st_drop_geometry() |>
  arrange(desc(total_area_depth)) |>
  slice_head(n = 10)
```

## 3. Exploratory hierarchical clustering

The following chunk uses the dong-level flood indicators to run an exploratory hierarchical clustering analysis. All indicators are standardized before calculating distances because they are measured on different scales. The clustering result is not yet interpreted or mapped here; cluster profiles and spatial visualization will be handled in the next step.

```{r}
#| label: hierarchical-clustering-flood-indicators
#| message: false
#| warning: false
#| fig-width: 10
#| fig-height: 7

cluster_indicators <- c(
  "log_total_damage_count",
  "log_total_F_AREA",
  "log_total_area_depth",
  "n_damage_years",
  "area_depth_evenness",
  "max_year_share_area_depth"
)

dong_cluster_input <- dong_flood_indicators |>
  st_drop_geometry() |>
  select(ADM_CD_2025, ADM_NM, all_of(cluster_indicators)) |>
  mutate(
    across(
      all_of(cluster_indicators),
      \(x) coalesce(as.numeric(x), 0)
    )
  )

dong_cluster_matrix <- dong_cluster_input |>
  select(all_of(cluster_indicators)) |>
  as.matrix()

rownames(dong_cluster_matrix) <- dong_cluster_input$ADM_CD_2025

# Standardize indicators before Euclidean distance and Ward.D2 clustering.
dong_cluster_scaled <- scale(dong_cluster_matrix)
dong_cluster_distance <- dist(dong_cluster_scaled, method = "euclidean")
dong_flood_hclust <- hclust(dong_cluster_distance, method = "ward.D2")

# Exploratory dendrogram. Change candidate_k to inspect a specific cluster cut.
candidate_k <- 4

plot(
  dong_flood_hclust,
  labels = FALSE,
  hang = -1,
  main = "Hierarchical clustering of flood indicators",
  xlab = "2025 administrative dongs",
  ylab = "Ward.D2 height"
)
rect.hclust(
  dong_flood_hclust,
  k = candidate_k,
  border = "red"
)

# Optional cluster membership vector for the current exploratory cut.
# This is not yet joined back to the sf object.
dong_hclust_group <- cutree(dong_flood_hclust, k = candidate_k)

# Compare several possible cluster cuts with rect.hclust.
plot_hclust_k <- function(hc, k) {
  dend_data <- ggdendro::dendro_data(hc, type = "rectangle")
  
  # Cut tree and identify x-ranges for each cluster
  cluster_df <- tibble::tibble(
    label = names(cutree(hc, k = k)),
    cluster = factor(cutree(hc, k = k))
  )
  
  label_df <- dend_data$labels |>
    dplyr::mutate(label = as.character(label)) |>
    dplyr::left_join(cluster_df, by = "label")
  
  rect_df <- label_df |>
    dplyr::group_by(cluster) |>
    dplyr::summarise(
      xmin = min(x) - 0.5,
      xmax = max(x) + 0.5,
      .groups = "drop"
    ) |>
    dplyr::mutate(
      ymin = -Inf,
      ymax = max(dend_data$segments$yend, dend_data$segments$y, na.rm = TRUE) * 0.38
    )
  
  ggplot() +
    geom_segment(
      data = dend_data$segments,
      aes(x = x, y = y, xend = xend, yend = yend),
      linewidth = 0.25
    ) +
    geom_rect(
      data = rect_df,
      aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax),
      fill = NA,
      color = "red",
      linewidth = 0.35
    ) +
    labs(
      title = paste("k =", k),
      x = NULL,
      y = "Height"
    ) +
    theme_minimal(base_size = 10) +
    theme(
      axis.text.x = element_blank(),
      axis.ticks.x = element_blank(),
      panel.grid = element_blank(),
      plot.title = element_text(hjust = 0.5, face = "bold")
    )
}

hclust_plots <- purrr::map(3:6, ~ plot_hclust_k(dong_flood_hclust, .x))

(hclust_plots[[1]] + hclust_plots[[2]]) /
  (hclust_plots[[3]] + hclust_plots[[4]])
```

## 4. Cluster-specific flood-damage profiles

The following section fixes the number of clusters at `k = 5`, joins the cluster membership back to the 2025 administrative-dong layer, summarizes the six clustering indicators by cluster, and assigns interpretable labels based on the hypothesized flood-damage typology.

```{r}
#| label: cluster-profile-labeling
#| message: false
#| warning: false

final_k <- 5

dong_cluster_membership <- tibble::tibble(
  ADM_CD_2025 = names(cutree(dong_flood_hclust, k = final_k)),
  cluster_id = cutree(dong_flood_hclust, k = final_k)
) |>
  mutate(
    cluster_id = factor(cluster_id, levels = sort(unique(cluster_id)))
  )

dong_cluster_labeled <- dong_flood_indicators |>
  left_join(dong_cluster_membership, by = "ADM_CD_2025")

cluster_profile_for_labeling <- dong_cluster_labeled |>
  st_drop_geometry() |>
  group_by(cluster_id) |>
  summarise(
    n_dongs = n(),
    mean_log_total_damage_count = mean(log_total_damage_count, na.rm = TRUE),
    mean_log_total_F_AREA = mean(log_total_F_AREA, na.rm = TRUE),
    mean_log_total_area_depth = mean(log_total_area_depth, na.rm = TRUE),
    mean_n_damage_years = mean(n_damage_years, na.rm = TRUE),
    mean_area_depth_evenness = mean(area_depth_evenness, na.rm = TRUE),
    mean_max_year_share_area_depth = mean(max_year_share_area_depth, na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(
    raw_damage_index = rowMeans(
      pick(
        mean_log_total_damage_count,
        mean_log_total_F_AREA,
        mean_log_total_area_depth
      )
    ),
    raw_recurrence_index = rowMeans(
      pick(
        mean_n_damage_years,
        mean_area_depth_evenness
      )
    ),
    raw_concentration_index = mean_max_year_share_area_depth
  ) |>
  mutate(
    damage_index = as.numeric(scale(raw_damage_index)),
    recurrence_index = as.numeric(scale(raw_recurrence_index)),
    concentration_index = as.numeric(scale(raw_concentration_index))
  )

very_low_damage_cluster <- cluster_profile_for_labeling |>
  slice_min(damage_index, n = 1, with_ties = FALSE) |>
  pull(cluster_id)

candidate_clusters <- cluster_profile_for_labeling |>
  filter(cluster_id != very_low_damage_cluster)

high_recurrent_cluster <- candidate_clusters |>
  slice_max(
    damage_index + recurrence_index - 0.25 * concentration_index,
    n = 1,
    with_ties = FALSE
  ) |>
  pull(cluster_id)

high_concentrated_cluster <- candidate_clusters |>
  filter(cluster_id != high_recurrent_cluster) |>
  slice_max(
    damage_index + concentration_index - 0.25 * recurrence_index,
    n = 1,
    with_ties = FALSE
  ) |>
  pull(cluster_id)

moderate_recurrent_cluster <- candidate_clusters |>
  filter(!cluster_id %in% c(high_recurrent_cluster, high_concentrated_cluster)) |>
  slice_max(
    recurrence_index - 0.15 * damage_index,
    n = 1,
    with_ties = FALSE
  ) |>
  pull(cluster_id)

low_concentrated_cluster <- candidate_clusters |>
  filter(
    !cluster_id %in% c(
      high_recurrent_cluster,
      high_concentrated_cluster,
      moderate_recurrent_cluster
    )
  ) |>
  slice_max(
    concentration_index - damage_index,
    n = 1,
    with_ties = FALSE
  ) |>
  pull(cluster_id)

cluster_label_levels <- c(
  "1. High damage and recurrent",
  "2. High damage but concentrated",
  "3. Moderate or low damage but recurrent",
  "4. Low damage and episodic",
  "5. Very low or no reported damage"
)

cluster_label_lookup <- cluster_profile_for_labeling |>
  mutate(
    cluster_label = case_when(
      cluster_id == high_recurrent_cluster ~ cluster_label_levels[1],
      cluster_id == high_concentrated_cluster ~ cluster_label_levels[2],
      cluster_id == moderate_recurrent_cluster ~ cluster_label_levels[3],
      cluster_id == low_concentrated_cluster ~ cluster_label_levels[4],
      cluster_id == very_low_damage_cluster ~ cluster_label_levels[5],
      TRUE ~ "Unassigned flood-damage profile"
    )
  ) |>
  arrange(cluster_id) |>
  select(cluster_id, cluster_label)

dong_cluster_labeled <- dong_cluster_labeled |>
  left_join(cluster_label_lookup, by = "cluster_id") |>
  mutate(
    cluster_label = factor(cluster_label, levels = cluster_label_levels),
    cluster_number_label = paste0("Cluster ", cluster_id, ": ", cluster_label)
  )

cluster_profile_table <- dong_cluster_labeled |>
  st_drop_geometry() |>
  group_by(cluster_id, cluster_label) |>
  summarise(
    n_dongs = n(),
    across(
      all_of(cluster_indicators),
      \(x) round(mean(x, na.rm = TRUE), 3),
      .names = "{.col}"
    ),
    .groups = "drop"
  ) |>
  arrange(cluster_id)

knitr::kable(
  cluster_profile_table,
  caption = "Mean values of the six flood-damage indicators by final k = 5 cluster."
)
```

```{r}
#| label: cluster-profile-boxplot
#| message: false
#| warning: false
#| fig-width: 11
#| fig-height: 8

cluster_map_palette <- setNames(
  c("#b2182b", "#ef8a62", "#2166ac", "#67a9cf", "#d9d9d9"),
  cluster_label_levels
)

dong_cluster_scaled_df <- as.data.frame(dong_cluster_scaled) |>
  tibble::rownames_to_column("ADM_CD_2025")

cluster_indicator_scaled_long <- dong_cluster_labeled |>
  st_drop_geometry() |>
  select(ADM_CD_2025, cluster_id, cluster_label) |>
  left_join(dong_cluster_scaled_df, by = "ADM_CD_2025") |>
  pivot_longer(
    cols = all_of(cluster_indicators),
    names_to = "indicator",
    values_to = "z_score"
  )

ggplot(
  cluster_indicator_scaled_long,
  aes(x = cluster_label, y = z_score, fill = cluster_label)
) +
  geom_hline(yintercept = 0, color = "grey55", linewidth = 0.25) +
  geom_boxplot(
    width = 0.65,
    outlier.alpha = 0.35,
    linewidth = 0.25
  ) +
  facet_wrap(~ indicator, ncol = 2) +
  scale_fill_manual(values = cluster_map_palette) +
  coord_flip() +
  labs(
    title = "k = 5 cluster profiles across standardized flood-damage indicators",
    x = NULL,
    y = "Standardized indicator value"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    legend.position = "none",
    panel.grid.minor = element_blank(),
    strip.text = element_text(face = "bold")
  )
```

## 5. Spatial visualization of the final clusters

The final `k = 5` cluster number and label are attached to each 2025 administrative dong. The map is drawn in EPSG:5179. Administrative-dong polygons are filled by cluster label with thin boundaries, while district boundaries are overlaid without fill and with thicker lines.

```{r}
#| label: final-cluster-tmap
#| message: false
#| warning: false
#| fig-width: 9
#| fig-height: 9

dong_cluster_map_5179 <- dong_cluster_labeled |>
  st_transform(5179)

seoul_sgg_5179 <- seoul_sgg |>
  st_transform(5179)

tmap_mode("plot")

tm_shape(dong_cluster_map_5179) +
  tm_polygons(
    col = "cluster_label",
    palette = cluster_map_palette,
    title = "Flood-damage cluster",
    border.col = "grey70",
    lwd = 0.15
  ) +
  tm_shape(seoul_sgg_5179) +
  tm_borders(
    col = "grey20",
    lwd = 1.2
  ) +
  tm_layout(
    title = "Administrative-dong flood-damage clusters, Seoul (k = 5)",
    legend.outside = TRUE,
    frame = FALSE
  )

flood_risk <- st_read(
  file.path(data_dir, "침수예상도/DS_Flooding_1_5179.shp"),
  quiet = TRUE
) |>
  st_make_valid()

tm_shape(dong_cluster_map_5179) +
  tm_polygons(
    col = "cluster_label",
    palette = cluster_map_palette,
    title = "Flood-damage cluster",
    border.col = "grey70",
    lwd = 0.15
  ) +
  tm_shape(seoul_sgg_5179) +
  tm_borders(
    col = "grey20",
    lwd = 1.2
  ) +
  tm_shape(flood_risk) +
  tm_borders(
    col = "grey20",
    lwd = 0.5
  ) +
  tm_layout(
    title = "Administrative-dong flood-damage clusters, Seoul (k = 5)",
    legend.outside = TRUE,
    frame = FALSE
  )
```

```{r}
#| message: false
#| warning: false
flood_analysis <- st_read(
  file.path(data_dir, "/final/seoul_flood_analysis_final.gpkg"),
  quiet = TRUE
) |>
  st_make_valid()

#------------------------------------------------------------
# 1. Prepare model data
#------------------------------------------------------------

model_data <- flood_analysis |>
  st_drop_geometry() |>
  mutate(
    Magnitude = factor(Magnitude, levels = c("LOW", "HIGH")),
    Repeatability = factor(Repeatability, levels = c("EPISODIC", "RECURRENT"))
  )

#------------------------------------------------------------
# 2. Magnitude model: HIGH vs LOW
#------------------------------------------------------------

magnitude_data <- model_data |>
  select(
    Magnitude,
    log_AtRisk,
    GM_Rate,
    SINGLE_HH_,
    Frgn_HH_R,
    NonHous_R,
    Youth_R,
    Elderly_R
  ) |>
  na.omit()

model_magnitude <- glm(
  Magnitude ~ log_AtRisk + GM_Rate + Frgn_HH_R +
    NonHous_R + Youth_R + SINGLE_HH_:Elderly_R,
  data = magnitude_data,
  family = binomial(link = "logit")
)

summary(model_magnitude)

magnitude_or <- broom::tidy(
  model_magnitude,
  exponentiate = TRUE,
  conf.int = TRUE
)

magnitude_or
```

```{r}
#| message: false
#| warning: false
#------------------------------------------------------------
# 3. Repeatability model: RECURRENT vs EPISODIC
#    Exclude no-damage cluster, assumed to be cluster 5
#------------------------------------------------------------

repeatability_data <- model_data |>
  filter(cluster_id == 1 | cluster_id == 2) |>
  select(
    area_depth_evenness,
    Repeatability,
    log_AtRisk,
    GM_Rate,
    SINGLE_HH_,
    Frgn_HH_R,
    NonHous_R,
    Youth_R,
    Elderly_R
  ) |>
  na.omit()

model_repeatability <- glm(
  Repeatability ~ GM_Rate*SINGLE_HH_ + Frgn_HH_R +
    NonHous_R + Elderly_R,
  family = binomial(link = "logit"),
  data = repeatability_data
)

summary(model_repeatability)

repeatability_or <- broom::tidy(
  model_repeatability,
  exponentiate = TRUE,
  conf.int = TRUE
)

repeatability_or
```

```{r}
#| message: false
#| warning: false


#------------------------------------------------------------
# 1. Prepare data
#------------------------------------------------------------

multi_data <- flood_analysis |>
  filter(cluster_id != 5) |>
  st_drop_geometry() |>
  mutate(
    # Proxy for potential elderly single-person household concentration
    single_elderly_proxy = SINGLE_HH_ * Elderly_R,

    # Set reference category
    cluster_label = factor(
      cluster_label,
      levels = c(
        "4. Low damage and episodic",
        "1. High damage and recurrent",
        "2. High damage but concentrated",
        "3. Moderate or low damage but recurrent"
      )
    )
  ) |>
  select(
    cluster_label,
    log_AtRisk,
    GM_Rate,
    SINGLE_HH_,
    Frgn_HH_R,
    NonHous_R,
    Youth_R,
    Elderly_R,
    single_elderly_proxy
  ) |>
  na.omit()

table(multi_data$cluster_label)

#------------------------------------------------------------
# 2. Multinomial logistic regression
#------------------------------------------------------------

model_multi <- nnet::multinom(
  cluster_label ~ log_AtRisk + GM_Rate + Frgn_HH_R + NonHous_R +
    Youth_R + SINGLE_HH_:Elderly_R + GM_Rate:Elderly_R,
  data = multi_data,
  trace = FALSE
)

multi_or <- broom::tidy(
  model_multi,
  exponentiate = TRUE,
  conf.int = TRUE
)

multi_or_table <- multi_or |>
  filter(term != "(Intercept)") |>
  mutate(
    outcome = y.level,
    predictor = case_when(
      term == "log_AtRisk" ~ "Log at-risk households",
      term == "GM_Rate" ~ "Gross migration rate",
      term == "Frgn_HH_R" ~ "Foreign household share",
      term == "NonHous_R" ~ "Non-housing residence share",
      term == "Youth_R" ~ "Youth population share",
      term == "single_elderly_proxy" ~ "Single elderly proxy",
      TRUE ~ term
    ),
    p_value = case_when(
      p.value < 0.001 ~ "<0.001",
      TRUE ~ sprintf("%.3f", p.value)
    ),
    sig = case_when(
      p.value < 0.001 ~ "***",
      p.value < 0.01 ~ "**",
      p.value < 0.05 ~ "*",
      p.value < 0.1 ~ "†",
      TRUE ~ ""
    ),
    `Odds ratio` = paste0(sprintf("%.2f", estimate), sig),
    `95% CI` = paste0("[", sprintf("%.2f", conf.low), ", ", sprintf("%.2f", conf.high), "]")
  ) |>
  select(
    Outcome = outcome,
    Predictor = predictor,
    `Odds ratio`,
    `95% CI`,
    `p-value` = p_value
  )

multi_or_table |>
  kable(
    format = "html",
    caption = "Multinomial logistic regression results predicting flood-damage cluster membership",
    align = c("l", "l", "r", "r", "r"),
    escape = FALSE
  ) |>
  kable_styling(
    full_width = FALSE,
    bootstrap_options = c("striped", "hover", "condensed"),
    position = "center",
    font_size = 13
  ) |>
  collapse_rows(
    columns = 1,
    valign = "top"
  ) |>
  footnote(
    general = "Reference category is the first level of the outcome factor. Values are odds ratios. 95% confidence intervals are shown in brackets. † p < 0.10, * p < 0.05, ** p < 0.01, *** p < 0.001.",
    general_title = "Note: ",
    footnote_as_chunk = TRUE
  )
```

```{r}
multi_null <- nnet::multinom(
  cluster_label ~ 1,
  data = multi_data,
  trace = FALSE
)

multi_full <- nnet::multinom(
  cluster_label ~ log_AtRisk + GM_Rate + Frgn_HH_R + NonHous_R +
    Youth_R + single_elderly_proxy +
    log_AtRisk:GM_Rate +
    GM_Rate:NonHous_R +
    log_AtRisk:NonHous_R,
  data = multi_data,
  trace = FALSE
)

multi_step <- MASS::stepAIC(
  multi_null,
  scope = list(
    lower = multi_null,
    upper = multi_full
  ),
  direction = "both",
  trace = TRUE
)

summary(multi_step)
AIC(multi_step)
formula(multi_step)

multi_step_or <- broom::tidy(
  multi_step,
  exponentiate = TRUE,
  conf.int = TRUE
)

multi_step_or
```
